---
title: "lab-03"
format: html
---

```{r}
library(tidyverse)
library(zoo)
library(flextable)
library(readr)
library(patchwork)
library(lubridate)
library(ggplot2)
library(dplyr)
```

Having open data is so important, because allowing the public to research data on their own is extremely important for scientific progress. Knowledge is power for preventing future disasters such as the COVID-19 pandemic, so when this data becomes inaccessible, we lose our ability to respond to future threats.

```{r}

data = read_csv('https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv')
my.date <- as.Date("2022-02-01")
my.state <- "Colorado"

library(dplyr)

colorado_data <- data %>%
  filter(state == my.state) %>%
  group_by(county) %>%
  arrange(county, date) %>%
  mutate(
    new_cases = cases - lag(cases, default = 0),
    new_deaths = deaths - lag(deaths, default = 0)
  ) %>%
  ungroup() %>%
  mutate(FIPS = str_pad(fips, width = 5, side = "left", pad = "0"))

top_cumulative_cases <- colorado_data %>%
  filter(date == my.date) %>%
  arrange(desc(cases)) %>%
  select(county, cases) %>%
  slice_head(n = 5)

top_cumulative_cases

top_new_cases <- colorado_data %>%
  filter(date == my.date) %>%
  arrange(desc(new_cases)) %>%
  select(county, new_cases) %>%
  slice_head(n = 5)

top_new_cases


```

```{r}
pop_url <- 'https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/totals/co-est2023-alldata.csv'
pop_data <- read_csv(pop_url)
library(dplyr)

pop_clean <- pop_data %>%
  mutate(
    STATE = sprintf("%02d", as.numeric(STATE)),
    COUNTY = sprintf("%03d", as.numeric(COUNTY)),
    FIPS = paste0(STATE, COUNTY)
  
 ) %>%
  filter(COUNTY != "000") %>%
  select(FIPS, contains("NAME"), contains("2021"))
```

```{r}
names(pop_clean)  
dim(pop_clean)     
str(pop_clean)    
summary(pop_clean) 
range(pop_clean$POPESTIMATE2021)  


```

```{r}

covid_joined <- colorado_data %>%
  left_join(pop_clean, by = "FIPS") %>%
  mutate(
    cases_per_100k = (cases / POPESTIMATE2021) * 100000,
    new_cases_per_100k = (new_cases / POPESTIMATE2021) * 100000,
    new_deaths_per_100k = (new_deaths / POPESTIMATE2021) * 100000
  )


```

```{r}

target_date <- as.Date("2021-01-01")
top_cumulative_per_capita <- covid_joined %>%
  filter(date == target_date) %>%
  arrange(desc(cases_per_100k)) %>%
  select(county, cases, POPESTIMATE2021, cases_per_100k) %>%
  slice_head(n = 5) %>%
  flextable() %>%
  set_caption("Top 5 Colorado Counties by Cumulative COVID-19 Cases per 100,000 (2021-01-01)")


top_new_per_capita <- covid_joined %>%
  filter(date == target_date) %>%
  arrange(desc(new_cases_per_100k)) %>%
  select(county, new_cases, POPESTIMATE2021, new_cases_per_100k) %>%
  slice_head(n = 5) %>%
  flextable() %>%
  set_caption("Top 5 Colorado Counties by New COVID-19 Cases per 100,000 (2021-01-01)")


top_cumulative_per_capita
top_new_per_capita


```

```{r}


latest_date <- max(covid_joined$date, na.rm = TRUE)


last_14_days <- covid_joined %>%
  filter(date > (latest_date - 14))


rolling_summary <- last_14_days %>%
  group_by(county) %>%
  summarise(
    pop_2021 = first(POPESTIMATE2021),
    total_new_cases = sum(new_cases, na.rm = TRUE),
    cases_per_100k_14day = (total_new_cases / pop_2021) * 100000,
    .groups = "drop"
  )


top_rolling <- rolling_summary %>%
  arrange(desc(cases_per_100k_14day)) %>%
  slice_head(n = 5) %>%
  flextable::flextable() %>%
  flextable::set_caption("Top 5 Colorado Counties by New Cases per 100,000 (Last 14 Days)")
 
watch_list_count <- rolling_summary %>%
  filter(cases_per_100k_14day > 100) %>%
  nrow()


top_rolling
watch_list_count



```

```{r}



covid_2021 <- covid_joined %>%
  filter(year(date) == 2021)

covid_deaths_2021 <- covid_2021 %>%
  group_by(county) %>%
  summarise(
    covid_deaths = sum(new_deaths, na.rm = TRUE),
    .groups = "drop"
  )

pop_deaths <- pop_clean %>%
  rename(county = CTYNAME) %>%
  select(county, total_deaths = DEATHS2021)  # make sure DEATHS2021 exists


death_analysis <- covid_deaths_2021 %>%
  left_join(pop_deaths, by = "county") %>%
  filter(!is.na(total_deaths), total_deaths > 0) %>%
  mutate(
    pct_covid_deaths = 100 * covid_deaths / total_deaths
  )

high_pct_deaths <- death_analysis %>%
  filter(pct_covid_deaths >= 20)

ggplot(high_pct_deaths, aes(x = reorder(county, pct_covid_deaths), y = pct_covid_deaths)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(
    title = "Counties Where ≥ 20% of 2021 Deaths Were from COVID",
    x = "County",
    y = "Percent of Deaths from COVID"
  ) +
  theme_minimal()


```

```{r}

target_states <- c("New York", "Colorado", "Alabama", "Ohio")

state_daily <- data %>%
  filter(state %in% c("New York", "Colorado", "Alabama", "Ohio")) %>%
  group_by(state, date) %>%
  summarise(
    daily_cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(state) %>%
  arrange(date) %>%
  mutate(
    new_cases = daily_cases - lag(daily_cases),
    roll_7 = rollmean(new_cases, k = 7, fill = NA, align = "right")
  ) %>%
  ungroup()
ggplot(state_daily, aes(x = date)) +
  geom_col(aes(y = new_cases), fill = "lightgray") +
  geom_line(aes(y = roll_7), color = "firebrick", size = 1) +
  facet_wrap(~state, scales = "free_y") +
  labs(
    title = "Daily New COVID-19 Cases and 7-Day Rolling Average",
    y = "New Cases",
    x = "Date"
  ) +
  theme_minimal()
state_pop <- tibble(
  state = c("New York", "Colorado", "Alabama", "Ohio"),
  pop_2021 = c(20201249, 5773714, 5024279, 11799448)
)

state_daily <- state_daily %>%
  left_join(state_pop, by = "state") %>%
  mutate(
    new_cases_per_100k = (new_cases / pop_2021) * 100000,
    roll_7_per_100k = (roll_7 / pop_2021) * 100000
  )

ggplot(state_daily, aes(x = date, y = roll_7_per_100k, color = state)) +
  geom_line(size = 1.2) +
  labs(
    title = "7-Day Rolling Average of New COVID Cases per 100,000",
    x = "Date",
    y = "New Cases per 100,000 Residents"
  ) +
  theme_minimal() +
  theme(legend.title = element_blank())



```

Scaling the populations makes it look worse for the lower-population states and better for the higher-population states. The real value in this scale is that we can more apparently see trends in the data, for example, we can tell that alabama had a more considerable spike right before the main spike in 2022.

```{r}
centroids_url <- "https://raw.githubusercontent.com/mikejohnson51/csu-ess-330/refs/heads/main/resources/county-centroids.csv"
county_centroids <- readr::read_csv(centroids_url)

covid_with_coords <- data %>%
  filter(!is.na(fips)) %>%
  left_join(county_centroids, by = "fips")
wmc_daily <- covid_with_coords %>%
  group_by(date) %>%
  summarise(
    total_cases = sum(cases, na.rm = TRUE),
    WMC_LON = weighted.mean(LON, cases, na.rm = TRUE),
    WMC_LAT = weighted.mean(LAT, cases, na.rm = TRUE),
    month = format(date, "%m"),
    .groups = "drop"
  )
ggplot() +
  borders("state", fill = "gray90", colour = "white") +
  geom_point(data = wmc_daily, aes(x = WMC_LON, y = WMC_LAT, size = total_cases, color = month), alpha = 0.7) +
  scale_size_continuous(range = c(1, 8)) +
  labs(
    title = "Weighted Mean Center of U.S. COVID-19 Cases Over Time",
    x = "Longitude",
    y = "Latitude",
    size = "Total Cases",
    color = "Month"
  ) +
  theme_minimal()

```

This chart would suggest that the cases began clustered in the east, and travelled west.

```{r}

wmc_centers <- covid_with_coords %>%
  filter(!is.na(cases), !is.na(deaths)) %>%
  group_by(date) %>%
  summarise(
    total_cases = sum(cases, na.rm = TRUE),
    total_deaths = sum(deaths, na.rm = TRUE),
    WMC_LON_cases = weighted.mean(LON, cases, na.rm = TRUE),
    WMC_LAT_cases = weighted.mean(LAT, cases, na.rm = TRUE),
    WMC_LON_deaths = weighted.mean(LON, deaths, na.rm = TRUE),
    WMC_LAT_deaths = weighted.mean(LAT, deaths, na.rm = TRUE),
    month = format(date, "%m"),
    .groups = "drop"
  )
plot_cases <- ggplot() +
  borders("state", fill = "gray90", colour = "white") +
  geom_point(data = wmc_centers, aes(x = WMC_LON_cases, y = WMC_LAT_cases, size = total_cases, color = as.numeric(month)), alpha = 0.7) +
  scale_size_continuous(range = c(1, 8)) +
  scale_color_viridis_c(option = "A") +
  labs(
    title = "Weighted Mean Center of COVID Cases",
    x = "Longitude", y = "Latitude", size = "Total Cases", color = "Month (1–12)"
  ) +
  theme_minimal()

plot_deaths <- ggplot() +
  borders("state", fill = "gray90", colour = "white") +
  geom_point(data = wmc_centers, aes(x = WMC_LON_deaths, y = WMC_LAT_deaths, size = total_deaths, color = as.numeric(month)), alpha = 0.7) +
  scale_size_continuous(range = c(1, 8)) +
  scale_color_viridis_c(option = "C") +
  labs(
    title = "Weighted Mean Center of COVID Deaths",
    x = "Longitude", y = "Latitude", size = "Total Deaths", color = "Month (1–12)"
  ) +
  theme_minimal()

plot_cases + plot_deaths


```
